---
title: "Fatal US School Shootings Are More Likely in Targeted and Multi-Victim Events: Evidence from Logistic Regression on Contextual Factors"
author: "Amanda Ng"
date: "2025-10-25"
output: pdf_document
---

```{r, include=FALSE}
# Load packages
library(tidyr)
library(dplyr)
library(arrow)
library(knitr)
library(ggplot2)
library(car)      
library(broom)
library(pROC)

# Load data
shootings = read_parquet("shootings_cleaned.parquet")
```

# Abstract

# 1. Introduction

Gun violence in schools remains one of the most distressing forms of firearm-related harm in the United States (citation). Each incident causes immediate loss and trauma while also increasing public concern over the safety of educational environments. Despite ongoing research, relatively little is known about why some school shootings result in fatalities while others do not. Understanding these factors is important for informing prevention strategies and school safety policies. Previous studies have examined individual-level factors such as shooter motivation, access to firearms, and mental health (citation). Fewer studies have examined contextual-level characteristics—features of the incident and school environment that may affect whether a shooting results in a fatality. This study examines these contextual factors to identify conditions that are associated with higher likelihoods of fatal outcomes.

The analysis draws on The Washington Post School Shootings Database, which documents every act of gunfire on U.S. primary and secondary school campuses during school hours since 1999. The estimand of interest is the probability that a school shooting results in at least one fatality. Using logistic regression, we modeled this probability as a function of several contextual characteristics, including whether the incident involved any injuries, the shooting type, the shooter’s demographics and relationship to the school community, and the broader racial and socioeconomic composition of the school.

Results show that shootings involving at least one injury, those classified as targeted attacks, and those committed out by shooters with no apparent or direct relationship to the school were more likely to result in fatalities. Shootings by individuals with no direct relationship to the school appear less predictable and may involve higher intent or premeditation. As such incidents are outside the reach of internal monitoring systems focused on students and staff, these findings emphasize the importance of external security measures, including controlled access to school facilities, improved information sharing with law enforcement and between schools, and preparedness for emergency response. Examining these patterns contributes to understanding the factors associated with school shooting fatality and can support the development of measures to reduce harm to students and school communities.

Section 2: Data describes the data source, alongside the cleaning and 
preparation of the dataset for analysis. A summary of key variables and exploratory data 
visualizations is also presented. Section 3: Modelling explains the logistic regression approach
used to analyze the relationship between fatality likelihood and contextual characteristics, 
along with justifications for the model design. Section 4: Results & Discussion presents 
the results and interprets the effects of predictors, as well as limitations of the study, including data quality and practical applicability of the models. Section 5: Conclusion summarizes the main findings and outlines directions for future studies. Supporting information, including model assumptions and diagnostic checks, is provided in Section 6: Appendix.

# 2. Data

## 2.1 Data source

The study draws on The Washington Post School Shootings Database (citation), a systematically compiled record of firearm incidents occurring at U.S. primary and secondary schools during school hours since the Columbine High School shooting on April 20, 1999. The database was developed by integrating multiple information channels, including Nexis archives, news reports, open-source databases, law enforcement records, school websites, and direct communications with school and police officials. Incidents were included only if gunfire occurred on campus immediately before, during, or after classes, thereby excluding after-hours events, accidental discharges injuring only the shooter, and suicides that posed no threat to others. Since no federal agency tracks school shootings, The Washington Post database represents one of the most comprehensive and systematically verified datasets available for studying these events. It further incorporates school-level demographic and enrollment information from the U.S. Department of Education’s Common Core of Data and Private School Universe Survey, allowing for contextual analyses of how school characteristics relate to the severity of shooting incidents.

## 2.2 Data overview

The raw data consists of 428 recorded school shooting incidents in the United States, spanning the period from April 20, 1999 onward. Each observation represents a distinct shooting event, and the dataset contains 50 variables, including identifiers for the school and district, temporal and geographic information, shooter demographic characteristics, shooting outcomes (number of killed and injured), school enrollment and students' demographics, and other factors such as the presence of a school resource officer. The dataset also includes detailed information on casualties, shooting type, and weapons involved. Variables of primary interest for this study include indicators of fatalities and injuries, shooting type, the shooter’s relationship to the school, demographic characteristics of the school population, and other contextual features that may influence the severity of the incident.

## 2.21 Data cleaning


R version and packages (citation)

## 2.22 Data description

```{r,include=FALSE}
# Compute raw data dimensions
nrow(shootings) # Total = 427
ncol(shootings) # Total = 9
```

The cleaned data consists of 427 observations and 9 variables. Below is a summary description of all variables in the cleeaned dataset.

```{r, echo=FALSE}
# Create a data frame
data_description <- data.frame(
  Variable = c(
    "Killing",
    "Injured",
    "School Type",
    "Shooting Type",
    "Shooter Age",
    "Shooter Gender",
    "Shooter Relationship",
    "Non-white Proportion",
    "Lunch Proportion"
  ),
  Type  = c(
    "Binary",
    "Binary ",
    "Categorical ",
    "Categorical ",
    "Numerical",
    "Categorical ",
    "Categorical",
    "Numerical ",
    "Numerical"
  ),
  Categories_or_Range = c(
    "1 = At least one killing (excluding shooter) occured",
    "1 = At least one injured (excluding shooter) occured",
    "Public, Private",
    "Indiscriminate, Targeted, Accidental, Suicide, Unclear",
    "6 to 74",
    "Male, Female",
    "Current Student, Former Student, Family/Guardian of Student, Non-Security Staff, Police/Security, Other/Unknown/No Connection",
    "0 to 1",
    "0 to 1"
  )
)

# Print the table using kable
kable(data_description, format = "markdown", col.names = c("Variable", "Type", "Categories/Range"))

```


## 2.3 Data summary and visualizations


# 5. Conclusion
These incidents may be less predictable because they occur outside existing social or disciplinary networks — schools can’t rely on familiar early-warning signs like student behavioral issues, conflict reports, or peer concerns.
Such shooters may enter the school with higher intent or preparation, possibly bringing more lethal weapons or planning attacks more deliberately.Because they are external actors, existing school-based prevention systems (focused on students and staff) may be less effective in identifying or deterring them beforehand.
schools canit easily prevent these attacks directly — but it highlights a different area of vulnerability and thus a different kind of prevention need:It suggests that school safety cannot rely solely on internal behavioral monitoring (e.g., identifying at-risk students).
There is a need for stronger external security coordination — for instance, improving access control, visitor screening, and information sharing between schools, law enforcement, and community organizations.
It also underscores the importance of rapid emergency response and preparedness training, since these events tend to be sudden and high-lethality when they occur.

# Appendix